Characterizing Optimizations to Memory Access Patterns using Architecture-Independent Program Features

### 2020-01-19 paper/paper.tex

\raggedbottom
\acmConference[IWOCL'20]{International Workshop on OpenCL}{April 27--29, 2020}{Munich, Germany}
%\editor{John B. Appleseed}
\title[Architecture-Independent Memory Access Pattern Analysis]{Characterising Optimizations to Memory Access Patterns using Architecture-Independent Program Features}

@@

	%This paper addresses the challenge of profiling a code base to help programmers effectively optimize their codes for high performance.
	HPC developers are faced with the challenge of optimizing OpenCL workloads for high performance on novel architectures.
	The Architecture-Independent Workload Characterization (AIWC) tool is a plugin for the Oclgrind OpenCL simulator that gathers metrics of OpenCL programs used to understand and predict program performance on an arbitrary given hardware architecture.
	However, AIWC metrics are not always easily interpreted and do not reflect some important memory access patterns affecting efficiency across architectures.
	We propose the concept of parallel spatial locality: the locality of memory accesses by OpenCL work-items (threads) executing simultaneously.
	We implement the parallel spatial locality metric in the AIWC framework and analyse gathered results on matrix multiply and the Extended OpenDwarfs OpenCL benchmarks.
	The differences in the parallel spatial locality metric between implementations of matrix multiply reflect the optimizations performed.
	The new metric showcases the memory access patterns in the OpenDwarfs benchmarks that affect their performance on various architectures.
	%Current methods of analysing program behaviour have largely relied on performance counters on various architectures to provide architecture dependent metrics of code behaviour. 
	%However, if workload characteristics are tied to architectural features specific to a particular system, these analysis frameworks may not generalize to alternative or future systems.
	%Considering the trend towards heterogeneous computing, we work towards a framework to guide developers in optimizing codes for any given hardware target using the architecture independent workload characterization (AIWC) tool -- specifically in profiling memory access patterns of code. 
	%Metrics must be easily interpreted and conceptually linked to patterns in hardware-specific memory access optimizations.
	%To this end we propose, implement and analyse the addition of two new metrics to AIWC and present a new approach to profile spatial locality intrinsic to parallel codes.
	The improvements suggested to AIWC will help HPC developers better understand memory access patterns of complex codes, a crucial step in guiding optimization of codes for arbitrary hardware targets.
	\end{abstract}

@@

	The OpenCL programming language is designed to support portability for heterogeneous systems by allowing a single code to execute on multiple hardware targets. 
	%It is possible for an engineer to write highly thread-divergent logical analysis code aimed at traditional CPUs; parallelisable data crunching kernels aimed at GPUs; and code to satisfy specialized computational requirements aimed at FPGAs -- all in a single code base using OpenCL. 
	The language provides support for individual \emph{kernels}, portions of parallelizable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}. This gives HPC software developers greater flexibility by allowing codes aimed at a range of hardware targets to be written in a single programming language environment.

@@

	\subsection{Optimising handling of shared memory}

@@

			\addplot[mark=*] coordinates {
				(simple,27.79)
				(coalA,16.36)
				(coalAB,17.02)
				(coalABT,16.34)
				(alignABT,30.21)
			};
			\addplot[mark=star] coordinates {
				(simple,319.71)
				(coalA,70.12)
				(coalAB,73.60)
				(coalABT,65.27)
				(alignABT,61.50)
			};	
			\end{axis}
			\end{tikzpicture} 
		}
		\subfloat{
			\begin{tikzpicture}
			\begin{axis}[symbolic x coords={simple,coalA,coalAB,coalABT,alignABT},
			xtick={simple,coalA,coalAB,coalABT,alignABT},
			xticklabel style={text height=2ex},
			ymin=0,
			ylabel={Execution time ($\mu s$)}, xlabel={Kernel},
			ylabel style={at={(axis description cs:0.1,.5)},anchor=south},
			legend entries={$N=1408$,$N=2048$},
			legend style={at={(0.6,0.45)},anchor=west}
			]

			\addplot[mark=triangle*] coordinates {
				(simple,46901.35)
				(coalA,8948.96)
				(coalAB,9237.5)
				(coalABT,5307)
				(alignABT,5272)
			};
			\addplot[mark=square] coordinates {
				(simple,130601.41)
				(coalA,28722.51)
				(coalAB,27169.72)
				(coalABT,16179.03)
				(alignABT,16071.90)
			};	
			\end{axis}
			\end{tikzpicture} 
		}
		\caption{NVIDIA Tesla P100 Runtime performance of section \ref{motivating example} codes averaged over 100 runs}
		\label{fig: matmul runtime}
	\end{figure*}

	Figure \ref{fig: matmul runtime} shows that across varying problem sizes, the vendor recommended optimizations to the matrix multiplication code lead to increased performance. Note that a matrix size of 80 is small -- overheads attached to optimization techniques may be causing the observed decrease in performance as some optimizations are performed.

@@

	\section{Discussion} \label{discussion}

	AIWC metrics are inherently lossy. Given the set of metrics calculated for an unknown kernel, we can not reconstruct the kernel based on these metrics. Specifically to memory access pattern analysis, the calculation of memory address entropy of a kernel relies only on the frequency distribution of memory accesses to all addresses accessed by the kernel. We observe that calculation of memory address entropy does not make use of temporal information. The order of sequential memory accesses performed by each thread, as well as addresses requested in parallel by multiple threads in an OpenCL thread group, are both vital in accurately characterising parallel codes.

@@

		\addplot[mark=*] coordinates {
			(0, 4.01)
			(1,3.51)
			(2,3.01)
			(3,2.50)
			(4,2.00)
			(5,2.00)
			(6,2.00)
			(7,2.00)
			(8,2.00)
			(9,1.5)
			(10,1.00)
		};

		\addplot[mark=star] coordinates {
			(0,4.24)
			(1,3.71)
			(2,3.18)
			(3,2.65)
			(4,2.12)
			(5,1.62)
			(6,1.12)
			(7,0.62)
			(8,0.12)
			(9,0.09)
			(10,0.06)
		};

		\addplot[mark=triangle*] coordinates {
			(0,4.45)
			(1,3.89)
			(2,3.33)
			(3,2.78)
			(4,2.23)
			(5,1.73)
			(6,1.23)
			(7,0.73)
			(8,0.23)
			(9,0.17)
			(10,0.11)
		};

		\addplot[mark=square] coordinates {
			(0,4.45)
			(1,3.45)
			(2,2.45)
			(3,1.45)
			(4,0.45)
			(5,0.40)
			(6,0.34)
			(7,0.28)
			(8,0.23)
			(9,0.17)
			(10,0.11)
		};

		\addplot[mark=o] coordinates {
			(0,4.45)
			(1,3.45)
			(2,2.45)
			(3,1.45)
			(4,0.45)
			(5,0.40)
			(6,0.34)
			(7,0.28)
			(8,0.23)
			(9,0.17)
			(10,0.11)
		};
		\end{axis}
		\end{tikzpicture}
		\caption{Calculated parallel spatial locality metric for matrix multiply kernels for 256 $\times$ 256 matrix multiplication}
		\label{fig: matmul parallel spatial locality}
	\end{figure*}

	There are three steps involved in generating an AIWC metric: \textit{recording}, \textit{calculating} and \textit{summarising} data collected from the symbolic execution of the kernel under inspection.

@@

	In figure \ref{fig: matmul parallel spatial locality} we see theoretical calculations for the proposed parallel spatial locality metric for each of the matrix multiplication kernels from section \ref{motivating example}. In particular we observe that the \texttt{coalescedABT} and \texttt{alignedABT} kernels have the steepest drop-offs in entropy as the number of bits skipped is increased, which correlates to better locality of parallel memory accesses. It is expected for these kernels to exhibit better parallel spatial locality than \texttt{simpleMultiply}, as \texttt{coalescedABT} and \texttt{alignedABT} make use of shared memory, where accesses tend to be localized simply due to the small size of shared on-chip memory typically available on GPUs. Further we find that the proposed metric successfully distinguishes between the \texttt{coalescedAB} and \texttt{coalescedABT} kernels. It accurately depicts a steeper drop-off for the more optimized \texttt{coalescedABT} kernel, where a larger proportion of parallel memory accesses make better use of the memory bank structure of GPU shared memory than all previous kernels. This is a significant improvement over the state-of-the-art AIWC metrics in characterising how codes localize shared memory accesses to better use the hardware provided.

	\section{Results}

	The Extended OpenDwarfs benchmark suite is a set of OpenCL workload classified according to the computation patterns or Dwarfs. First presented in 2006, the Dwarfs categorize 
	We present the results of running the AIWC tool on the Extended OpenDwarfs benchmark suite \cite{opendwarfs2017base,extendedopendwarfs}.

	% As all kernels access memory on word (32-bit) boundaries, there is no effect in dropping upto two bits in the PSL metric...
	\begin{figure*}[h!]
		\centering
		\begin{tikzpicture}
		\begin{axis}[symbolic x coords={0,1,2,3,4,5,6,7,8,9,10},
		width=0.7\textwidth,
		height=0.5\textwidth,
		xtick={0,1,2,3,4,5,6,7,8,9,10},
		xticklabel style={text height=2ex},
		ymin=0,
		ylabel={$n$-bits dropped Parallel Spatial Locality}, xlabel={\# of bits skipped},
		ylabel style={at={(axis description cs:0.05,.5)},anchor=south},
		legend entries={\texttt{GEM}, \texttt{NW*}, \texttt{SRAD Kernel 1},\texttt{SRAD Kernel 2},
			\texttt{CSR}, \texttt{BFS*}, \texttt{LUD Diagonal}, \texttt{LUD Internal}, \texttt{LUD Perimeter}, \texttt{TDM}},
		legend style={at={(0.55,0.75)},anchor=west}
		]

		\addplot[mark=*] coordinates {
			(0, 0.1139)
			(1, 0.1139)
			(2, 0.1139)
			(3, 0.1139)
			(4, 0.0989)
			(5, 0.08402)
			(6, 0.06914)
			(7, 0.05431)
			(8, 0.03938)
			(9, 0.02442)
			(10, 0.0124)
		};

		\addplot[mark=star] coordinates {
			(0, 2.175)
			(1, 2.175)
			(2, 2.175)
			(3, 2.175)
			(4, 2.175)
			(5, 2.175)
			(6, 2.175)
			(7, 2.175)
			(8, 2.175)
			(9, 2.175)
			(10, 2.175)
		};

		\addplot[mark=*] coordinates {
			(0, 7.232)
			(1, 7.232)
			(2, 7.232)
			(3, 7.232)
			(4, 6.351)
			(5, 5.461)
			(6, 4.570)
			(7, 3.678)
			(8, 3.648)
			(9, 3.633)
			(10, 3.625)
		};

		\addplot[mark=star] coordinates {
			(0, 7.098)
			(1, 7.098)
			(2, 7.098)
			(3, 7.098)
			(4, 6.223)
			(5, 5.344)
			(6, 4.464)
			(7, 3.583)
			(8, 3.567)
			(9, 3.559)
			(10, 3.556)
		};

		\addplot[mark=triangle*] coordinates {
			(0, 9.605)
			(1, 9.605)
			(2, 9.605)
			(3, 9.605)
			(4, 9.478)
			(5, 9.293)
			(6, 9.049)
			(7, 8.761)
			(8, 8.447)
			(9, 8.12)
			(10,7.233)
		};

		\addplot[mark=square] coordinates {
			(0,4.406)
			(1,4.406)
			(2,4.406)
			(3,4.406)
			(4,4.361)
			(5,4.273)
			(6,4.135)
			(7,3.941)
			(8,3.689)
			(9,3.382)
			(10,3.029)
		};

		\addplot[mark=o] coordinates {
			(0,1.548)
			(1,1.548)
			(2,1.548)
			(3,1.548)
			(4,1.305)
			(5,1.092)
			(6,0.9385)
			(7,0.8796)
			(8,0.8796)
			(9,0.8796)
			(10,0.8796)
		};
		\end{axis}
		\end{tikzpicture}
		\caption{Parallel spatial locality metric for selected OpenDwarfs benchmark kernels}
		\label{fig: opendwarfs parallel spatial locality}
	\end{figure*}
	\subsection{N-body methods: GEM}

@@

@@

	LUD in OpenDwarfs is a program to decompose an input $N\times N$ matrix as the product of one lower and one upper diagonal matrix. LUD is a Memory access patterns in a dense linear algebra workload such as LUD are typically highly regular and deterministic for each OpenCL work item, based on the the matrix dimension and offset parameters. The OpenDwarfs implementation of LUD \cite{opendwarfs2017base} splits the LUD computation into three kernels: LUD perimeter and diagonal kernels spawn work-items in a single work dimension while LUD internal performs work in both dimensions of the input matrix.

	LUD perimeter and diagonal partially benefit from memory access coalescing, for the lines of code where all the work-items access contiguous memory along a row of the input matrix. However, when threads parallelly access a column of the input matrix, multiple memory requests are made as addresses accessed are too distant to be coalesced into a single memory transaction on GPUs. This is reflected in the slow drop in parallel spatial locality for LUD perimeter and diagonal kernels.

	LUD internal %TODO: 2Dimensional work -- better parallelism

	%The use of creates imbalanced workload and divergent control flow within work-items in each kernel iteration. While the presence of conditional statements cause some thread divergence in the LUD diagonal kernel, the use of barriers following conditional statements synchronizes the threads. LUD diagonal and perimeter kernels , while LUD.

	...  %TODO: Divergence in the kernels. Make a diagram/code simplification for LUD.

@@

	Needleman-Wunsch is a dynamic programming algorithm used to perform protein sequence alignment by identifying the similarity level between two strings of Amino acids. Elements in a matrix are computed, with each element dependent on its west, north and north-west neighbours. This dependency enforces a wavefront computation pattern, travelling along the main diagonal of the matrix. Each iteration of the kernels computes over an antidiagonal of the matrix, starting at its top left corner and finishing at its bottom right corner.

	In particular, the wavefront pattern of computations causes each work-item to request distant memory addresses in a parallel fashion. Thus, parallel memory requests into the input matrix prohibit locality. We note the lack of any dropoff in parallel spatial locality as the number of bits dropped increases. This rightly indicates that memory addresses accessed at each logical timestamp are very distant. On GPUs this will translate to poor utilization of both memory access coalescing and caching. This trend in the parallel spatial locality metric suggests that a possible improvement for GPU performance would be to load blocks of the input matrix into on-chip local memory to reduce the number of global memory requests.

@@

	SRAD removes locally correlated noise from images by following a repeated grid update computation pattern. Conditional statements in the code cause thread divergence, potentially within a work-group, to handle boundary conditions. These boundaries constitute a small portion of the executed work-items, especially on larger data-sets and so the effective thread divergence is minimal. Thus, memory access patterns in SRAD are statically determined and relatively localized. The observed rapid drop-off in parallel spatial locality of the SRAD kernels (figure \ref{fig: opendwarfs parallel spatial locality}) is due to the highly localized nature of parallel memory accesses. At each point in the kernels' memory access profile, each work-item $(i,j)$ in an OpenCL work-group accesses the $(i,j)^{th}$ elements of various matrices. The sequential memory access pattern is non-linear since different matrices are accessed consecutively by the kernel, prohibiting ideal caching. However, memory requests made parallelly are highly localized, allowing memory access coalescing on GPUs, greatly reducing the number of memory requests made on each line of the kernel code on GPUs. Thus while cache hit rates are typically low on this benchmark, particularly on GPUs \cite{krommydas2016opendwarfs}, GPUs can hide the latency of global memory accesses through coalescing.

@@

	Graph traversal algorithms require pointer chasing operations to traverse nodes of a graph and perform calculations. The Breadth First Search implemented in OpenDwarfs calls OpenCL kernels to traverse nodes immediately connected to the list of nodes at each depth starting from the root node. BFS is characterized by an imbalanced workload per kernel launch depending on the degree of the nodes being operated on -- with only a proportion of launched nodes performing meaningful work. As such the AIWC features collected for each kernel invocation vary significantly. The parallel spatial locality metric collects the entropy of memory accesses at each time step, and workload imbalances are dealt with by averaging entropy scores across all the time steps in the execution of an OpenCL work-group.

>>>>

\raggedbottom
\acmConference[IWOCL'20]{International Workshop on OpenCL}{April 27--29, 2020}{Munich, Germany}
%\editor{John B. Appleseed}
\title[Architecture-Independent Memory Access Pattern Analysis]{Characterizing Optimizations to Memory Access Patterns using Architecture-Independent Program Features}

@@

	%This paper addresses the challenge of profiling a code base to help programmers effectively optimize their codes for high performance.
	HPC developers are faced with the challenge of optimizing OpenCL workloads for high performance on novel architectures.
	The Architecture Independent Workload Characterisation (AIWC) tool is a plugin for the Oclgrind OpenCL simulator that gathers metrics of OpenCL programs that can be used to understand and predict program performance on an arbitrary given hardware architecture.
	However, AIWC metrics are not always easily interpreted and do not reflect some important memory access patterns affecting efficiency across architectures.
	We propose the concept of parallel spatial locality -- the closeness of memory accesses simultaneously issued by OpenCL work-items (threads).
	We implement the parallel spatial locality metric in the AIWC framework, and analyse gathered results on matrix multiply and the Extended OpenDwarfs OpenCL benchmarks.
	The differences in the obtained parallel spatial locality metric across implementations of matrix multiply reflect the optimizations performed.
	The new metric can be used to distinguish between the OpenDwarfs benchmarks based on the memory access patters affecting their performance on various architectures.
	%Current methods of analysing program behaviour have largely relied on performance counters on various architectures to provide architecture dependent metrics of code behaviour. 
	%However, if workload characteristics are tied to architectural features specific to a particular system, these analysis frameworks may not generalize to alternative or future systems.
	%Considering the trend towards heterogeneous computing, we work towards a framework to guide developers in optimizing codes for any given hardware target using the architecture independent workload characterization (AIWC) tool -- specifically in profiling memory access patterns of code. 
	%Metrics must be easily interpreted and conceptually linked to patterns in hardware-specific memory access optimizations.
	%To this end we propose, implement and analyse the addition of two new metrics to AIWC and present a new approach to profile spatial locality intrinsic to parallel codes.
	The improvements suggested to AIWC will help HPC developers better understand memory access patterns of complex codes, a crucial step in guiding optimization of codes for arbitrary hardware targets.
	\end{abstract}

@@

	The OpenCL programming language is designed to support modern HPC software engineers in writing code that executes on multiple hardware targets. 
	%It is possible for an engineer to write highly thread-divergent logical analysis code aimed at traditional CPUs; parallelisable data crunching kernels aimed at GPUs; and code to satisfy specialized computational requirements aimed at FPGAs -- all in a single code base using OpenCL. 
	%The language provides support for individual \emph{kernels}, portions of parallelisable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}. 
	This gives HPC software developers greater flexibility by allowing codes aimed at a range of hardware targets to be written in a single programming language environment.

@@

	\subsection{Optimizing handling of shared memory}

@@

			\addplot coordinates {%[mark=*] coordinates {
				(simple,31)
				(coalA,19)
				(coalAB,20)
				(coalABT,18)
				(alignABT,18)
			};
			\addplot coordinates {%[mark=star] coordinates {
				(simple,325)
				(coalA,75)
				(coalAB,81)
				(coalABT,55)
				(alignABT,55)
			};	
			\end{axis}
			\end{tikzpicture} 
		}
		\subfloat{
			\begin{tikzpicture}
			\begin{axis}[symbolic x coords={simple,coalA,coalAB,coalABT,alignABT},
			xtick={simple,coalA,coalAB,coalABT,alignABT},
			xticklabel style={text height=2ex},
			ymin=0,
			ylabel={Execution time ($\mu s$)}, xlabel={Kernel},
			ylabel style={at={(axis description cs:0.1,.5)},anchor=south},
			legend entries={$N=1408$,$N=2048$},
			legend style={at={(0.6,0.45)},anchor=west}
			]

			\addplot [mark=*, color=brown] coordinates {%[mark=triangle*] coordinates {
				(simple,46920)
				(coalA,8783)
				(coalAB,9237.5)
				(coalABT,5307)
				(alignABT,5272)
			};
			\addplot [mark=star] coordinates {%[mark=square] coordinates {
				(simple,130682.0)
				(coalA,28602)
				(coalAB,27939)
				(coalABT,16179)
				(alignABT,16071)
			};	
			\end{axis}
			\end{tikzpicture} 
		}
		\caption{NVIDIA Tesla P100 Runtime performance of section \ref{motivating example} codes averaged over 100 runs}
		\label{fig: matmul runtime}
	\end{figure*}

	Figure \ref{fig: matmul runtime} shows that across varying problem sizes, the vendor recommended optimizations to the matrix multiplication code lead to increased performance. 
	%Note that a matrix size of 80 is small -- overheads attached to optimization techniques may be causing the observed decrease in performance as some optimizations are performed.

@@

	\section{Discussion } \label{discussion}

	AIWC metrics are inherently lossy. Given the set of metrics calculated for an unknown kernel, we can not reconstruct the kernel based on these metrics. Specifically to memory access pattern analysis, the calculation of memory address entropy of a kernel relies only on the frequency distribution of memory accesses to all addresses accessed by the kernel. We observe that calculation of memory address entropy does not make use of temporal information. The order of sequential memory accesses performed by each thread, as well as addresses requested in parallel by multiple threads in an OpenCL thread group, are both vital in accurately characterizing parallel codes.

@@

		\addplot coordinates {%[mark=*] coordinates {
			(0, 3.987)
			(1, 3.987)
			(2, 3.987)
			(3, 3.49)
			(4, 2.992)
			(5, 2.494)
			(6, 1.996)
			(7, 1.996)
			(8, 1.996)
			(9, 1.996)
			(10, 1.996)
		};

		\addplot coordinates {%[mark=star] coordinates {
			(0,3.952)
			(1,3.952)
			(2,3.952)
			(3,3.459)
			(4,2.965)
			(5,2.472)
			(6,1.979)
			(7,1.506)
			(8,1.033)
			(9,0.5598)
			(10,0.08691)
		};

		\addplot coordinates {%[mark=triangle*] coordinates {
			(0,4.145)
			(1,4.145)
			(2,4.145)
			(3,3.628)
			(4,3.11)
			(5,2.598)
			(6,2.075)
			(7,1.603)
			(8,1.13)
			(9,0.6579)
			(10,0.1854)
		};

		\addplot [mark=*, color=red] coordinates {%[mark=square] coordinates {
			(0,4.145)
			(1,4.145)
			(2,4.145)
			(3,3.2)
			(4,2.254)
			(5,1.308)
			(6,0.3621)
			(7,0.318)
			(8,0.2738)
			(9,0.2296)
			(10,0.1854)
		};

		\addplot coordinates {%[mark=o] coordinates {
			(0,4.145)
			(1,4.145)
			(2,4.145)
			(3,3.2)
			(4,2.254)
			(5,1.308)
			(6,0.3621)
			(7,0.318)
			(8,0.2738)
			(9,0.2296)	
			(10,0.1854)
		};
		\end{axis}
		\end{tikzpicture}
		\caption{Calculated parallel spatial locality metric for matrix multiply kernels for 256 $\times$ 256 matrix multiplication}
		\label{fig: matmul parallel spatial locality}
	\end{figure*}

	There are three steps involved in generating an AIWC metric: \textit{recording}, \textit{calculating} and \textit{summarizing} data collected from the symbolic execution of the kernel under inspection.

@@

	In figure \ref{fig: matmul parallel spatial locality} we see the proposed parallel spatial locality metric obtained using the AIWC tool for each of the matrix multiplication kernels from section \ref{motivating example}. In particular we observe that the \texttt{coalescedABT} and \texttt{alignedABT} kernels have the steepest drop-offs in entropy as the number of bits skipped is increased, which correlates to better locality of parallel memory accesses. It is expected for these kernels to exhibit better parallel spatial locality than \texttt{simpleMultiply}, as \texttt{coalescedABT} and \texttt{alignedABT} make use of shared memory, where accesses tend to be localized simply due to the small size of shared on-chip memory typically available on GPUs. Further we find that the proposed metric successfully distinguishes between the \texttt{coalescedAB} and \texttt{coalescedABT} kernels. It accurately depicts a steeper drop-off for the more optimized \texttt{coalescedABT} kernel, where a larger proportion of parallel memory accesses make better use of the memory bank structure of GPU shared memory than all previous kernels. This is a significant improvement over the state-of-the-art AIWC metrics in characterizing how codes localize shared memory accesses to better use the hardware provided.

	The Extended OpenDwarfs benchmark suite is a set of OpenCL workload classified according to the computation patterns or Dwarfs. 
	First presented in 2006, the 13 Berkeley Dwarfs are common computational and communication patterns that aim to capture the landscape of parallel computing workloads. 
	The Extended OpenDwarfs benchmark suite presents a set of highly diverse workloads that each fit the programming paradigm of one of the 13 Berkeley Dwarfs.
	The workloads presented are not optimized for any specific architecture -- hence optimizations using OpenCL local memory (which translates to CUDA shared memory) are not performed. 
	Many of these benchmarks can be run with up to four \emph{problem sizes} based on the sizes of caches found in modern CPU memory heirarchies \cite{johnston18opendwarfs}.
	We present the results of running the AIWC tool on selected benchmarks from the Extended OpenDwarfs benchmark suite \cite{opendwarfs2020head,johnston18opendwarfs} with a problem size setting of \emph{medium}, except for the GEM benchmark, which is run at a setting of \emph{tiny}.
	For benchmarks such as BFS with multiple kernel invocations per run, the AIWC parallel spatial locality metrics of the invocation with the highest executed LLVM-SPIR instructions is presented for analysis. Note that the presented benchmarks use 32 bit numbers (OpenCL \texttt{int} and \texttt{float} types), and so dropping up to two bits of the memory addresses accessed will not change the data entropy, since any addresses accessed are at least 4 bytes apart.

	%TODO: Add some info about EOD benchmarks here?

	% As all kernels access memory on word (32-bit) boundaries, there is no effect in dropping upto two bits in the PSL metric...
	\begin{figure*}[h!]
		\centering
		\begin{tikzpicture}
		\begin{axis}[symbolic x coords={0,1,2,3,4,5,6,7,8,9,10},
		width=0.7\textwidth,
		height=0.5\textwidth,
		xtick={0,1,2,3,4,5,6,7,8,9,10},
		xticklabel style={text height=2ex},
		ymin=0,
		ylabel={$n$-bits dropped Parallel Spatial Locality}, xlabel={\# of bits skipped},
		legend columns=4,
		ylabel style={at={(axis description cs:0.05,.5)},anchor=south},
		legend entries={\texttt{GEM}, \texttt{NW}, \texttt{SRAD Kernel 1},\texttt{SRAD Kernel 2},
			\texttt{CSR}, \texttt{BFS Kernel 1}, \texttt{BFS Kernel 2}, \texttt{LUD Diagonal}, \texttt{LUD Perimeter}, \texttt{LUD Internal}, \texttt{TDM}},
		legend style={at={(0.5,-0.15)},anchor=north}
		]

		\addplot coordinates {%[mark=*] coordinates {
			(0, 0.1139)
			(1, 0.1139)
			(2, 0.1139)
			(3, 0.0989)
			(4, 0.08402)
			(5, 0.06914)
			(6, 0.05431)
			(7, 0.03938)
			(8, 0.02442)
			(9, 0.0124)
			(10, 0.00038)
		};

		\addplot coordinates {%[mark=star] coordinates {
			(0, 2.175)
			(1, 2.175)
			(2, 2.175)
			(3, 2.175)
			(4, 2.175)
			(5, 2.175)
			(6, 2.175)
			(7, 2.175)
			(8, 2.175)
			(9, 2.175)
			(10, 2.175)
		};

		\addplot coordinates {%[mark=*] coordinates {
			(0, 7.176)
			(1, 7.176)
			(2, 7.176)
			(3, 6.301)
			(4, 5.418)
			(5, 4.533)
			(6, 3.648)
			(7, 3.619)
			(8, 3.605)
			(9, 3.598)
			(10, 3.594)
		};

		\addplot coordinates {%[mark=star] coordinates {
			(0, 7.097)
			(1, 7.097)
			(2, 7.097)
			(3, 6.223)
			(4, 5.344)
			(5, 4.464)
			(6, 3.583)
			(7, 3.567)
			(8, 3.559)
			(9, 3.554)
			(10, 3.553)
		};

		\addplot coordinates {%[mark=triangle*] coordinates {
			(0, 9.605)
			(1, 9.605)
			(2, 9.605)
			(3, 9.478)
			(4, 9.293)
			(5, 9.049)
			(6, 8.761)
			(7, 8.447)
			(8, 8.12)
			(9,7.233)
			(10,6.287)
		};

		\addplot coordinates { %[mark=square] coordinates {
			(0,4.933)	
			(1,4.933)
			(2,4.933)
			(3,4.89)
			(4,4.795)
			(5,4.652)
			(6,4.456)
			(7,4.215)
			(8,3.93)
			(9,3.608)
			(10,3.253)
		};

		\addplot coordinates { %[mark=square] coordinates {
			(0,4.691)
			(1,4.691)
			(2,4.691)
			(3,4.276)
			(4,3.760)
			(5,3.171)
			(6,2.552)
			(7,1.922)
			(8,1.286)
			(9,0.648)
			(10,0.00776)
		};

		\addplot coordinates { %[mark=o] coordinates {
			(0,1.548)
			(1,1.548)
			(2,1.548)
			(3,1.305)
			(4,1.092)
			(5,0.9385)
			(6,0.8796)
			(7,0.8796)
			(8,0.8796)
			(9,0.8796)
			(10,0.8796)
		};

	    \addplot coordinates { %[mark=star] coordinates {
	    	(0,3.803)
	    	(1,3.803)
	    	(2,3.803)
	    	(3,3.473)
	    	(4,3.144)
	    	(5,2.814)
	    	(6,2.485)
	    	(7,2.485)
	    	(8,2.485)
	    	(9,2.485)
	    	(10,2.485)
    	};

   		\addplot coordinates { %[mark=square*] coordinates {
   			(0,4.101)
   			(1,4.101)
   			(2,4.101)
   			(3,3.589)
   			(4,3.077)
   			(5,2.565)
   			(6,2.053)
   			(7,2.053)
   			(8,2.053)
   			(9,2.053)
   			(10,2.053)
   		};

		\end{axis}
		\end{tikzpicture}
		\caption{Parallel spatial locality metric for selected OpenDwarfs benchmark kernels}
		\label{fig: opendwarfs parallel spatial locality}
	\end{figure*}

	\subsection{N-body methods: GEM}

@@

	Performance results \cite{johnston18opendwarfs, krommydas2016opendwarfs} show that GPUs perform significantly better than CPUs for GEM, as memory unit stalls are at low levels for both CPUs and GPUs due to the highly efficient memory utilization of this benchmark. 
	As memory operations do not present a major slowdown, increased FPU compute capability of GPUs leads to far better performance on GPUs than CPUs \cite{krommydas2016opendwarfs}.

@@

	LUD in OpenDwarfs is a program to decompose an input $N\times N$ matrix as the product of one lower and one upper diagonal matrix. 
	Memory access patterns in a dense linear algebra workload such as LUD are typically highly regular and deterministic for each OpenCL work item, based on the matrix dimension and offset parameters. The OpenDwarfs implementation of LUD \cite{opendwarfs2020head} splits the LUD computation into three kernels: \texttt{LUD perimeter} and \texttt{LUD diagonal} kernels spawn work-items in a single work dimension; while \texttt{LUD internal} performs work in both dimensions of the input matrix.

	The LUD kernels partially benefit from memory access coalescing on GPUs, for the lines of code where all the work-items access contiguous memory along a row of the input matrix. 
	However, when threads parallelly access a column of the input matrix, multiple memory requests are made as addresses accessed are too distant to be coalesced into a single memory transaction on GPUs. 
	This is reflected in the drop in parallel spatial locality for the LUD kernels to approximately half the value at 0 bits skipped. Taking \texttt{LUD internal} as an example, the 0-bits skipped parallel spatial locality metric is 4.101, while the 10-bits skipped parallel spatial locality metric is 2.053. 
	The swift decline of the metric to 2.053 from 2 to 6-bits skipped parallel spatial locality indicates that approximately half the parallel memory accesses in \texttt{LUD internal} are highly localized -- where work-items simultaneously access contiguous memory along a matrix row. 
	Similar to the presented matrix multiply codes, LUD can be optimized for GPUs by effective utilization of shared memory \cite{opendwarfs2020head}.

	%The use of creates imbalanced workload and divergent control flow within work-items in each kernel iteration. While the presence of conditional statements cause some thread divergence in the LUD diagonal kernel, the use of barriers following conditional statements synchronizes the threads. LUD diagonal and perimeter kernels , while LUD.
	%TODO: Divergence in the kernels. Make a diagram/code simplification for LUD.

@@

	Needleman-Wunsch is a dynamic programming algorithm used to perform protein sequence alignment by identifying the similarity level between two strings of Amino acids. Elements in a matrix are computed, with each element dependent on its west, north and north-west neighbours. This dependency enforces a wavefront computation pattern, travelling along the main diagonal of the matrix. Each iteration of the kernels computes over an antidiagonal of the matrix, starting at its top left corner, and finishing at its bottom right corner.

	In particular, the wavefront pattern of computations causes each work-item to request distant memory addresses in a parallel fashion. Thus, parallel memory requests into the input matrix prohibit locality. We note the lack of any dropoff in parallel spatial locality as the number of bits dropped increases. This rightly indicates that memory addresses accessed at each logical timestamp are very distant. On GPUs this will translate to poor utilization of both memory access coalescing and caching \cite{krommydas2016opendwarfs}. This trend in the parallel spatial locality metric suggests that a possible improvement for GPU performance would be to load blocks of the input matrix into on-chip local memory to reduce the number of global memory requests -- this is typically performed in GPU optimized implementations of the Needleman-Wunsch algorithm \cite{opendwarfs2020head}.

@@

	SRAD removes locally correlated noise from images by following a repeated grid update computation pattern on image pixel grids. 
	Conditional statements in the code cause thread divergence, potentially within a work-group, to handle boundary conditions. 
	These boundaries constitute a small portion of the executed work-items, especially on larger data-sets and so the effective thread divergence is minimal.
	Memory access patterns in SRAD are statically determined and relatively localized. 
	Similar to the LUD and \texttt{simpleMultiply} kernels, both SRAD kernels observe a rapid drop-off in parallel spatial locality (figure \ref{fig: opendwarfs parallel spatial locality}) as the number of bits skipped is increased, with the metric stabilizing at approximately half its original value when 0 bits are skipped.

	At each point in the kernels' memory access profile, each work-item with global ID $(i,j)$ in an OpenCL work-group accesses the $(i,j)^{th}$ elements of various matrices. The sequential memory access pattern is non-linear since different matrices are accessed consecutively by the kernel, prohibiting ideal caching \cite{krommydas2016opendwarfs}. However, memory requests made simultaneously by a work-group always fall within a rectangular block of one of the matrices. 
	This allows memory access coalescing on GPUs for OpenCL work-items accessing contiguous matrix data, greatly reducing the number of memory requests made on each line of the kernel code on GPUs. 
	However, work-items accessing data along a column of the matrix do not observe memory access coalescing.
	Thus we observe that while cache hit rates are typically low on this benchmark, particularly on GPUs \cite{krommydas2016opendwarfs}, GPUs can hide the latency of global memory accesses through memory access coalescing to some extent.

@@

	Graph traversal algorithms require pointer chasing operations to traverse nodes of a graph and perform calculations. 
	The Breadth First Search implemented in OpenDwarfs calls two OpenCL kernels to traverse nodes immediately connected to the list of nodes at each depth starting from the root node. 
	BFS is characterized by an imbalanced workload per kernel launch depending on the degree of the nodes being operated on, with only a proportion of launched nodes performing meaningful work. 
	As such, the AIWC features collected for each kernel invocation vary significantly. 
	The parallel spatial locality metric collects the entropy of memory accesses at each time step, and workload imbalances across work-items are dealt with by averaging entropy scores across all the time steps in the execution of an OpenCL work-group.
